---
title: "ReproducibleResearchHW1"
author: "Nick Stuber"
date: "November 21, 2016"
output: html_document
---


```{r}
## Load in raw data into dataframe
rawData <- read.csv("activity.csv", 
                      sep = ',', 
                      header = TRUE,
                      na.strings = "NA")
```

# Q: What is mean total number of steps taken per day?

Removing rows with 'NA' values in the steps column
```{r}
updatedData <- rawData[!(is.na(rawData$steps)), ]
```

Sum of steps by the date
```{r}
stepsByDate <- aggregate(steps ~ date, updatedData, sum)
```

Create histogram of steps taken per day
```{r, echo=TRUE}
hist(stepsByDate$steps, 
     main = "Histogram of total number of steps per day",
     xlab = "Steps per day")
```

Calculating the mean of total number of steps taken per day
```{r}
mean(stepsByDate$steps)
```

Calculating the median of total number of steps taken per day
```{r}
median(stepsByDate$steps)
```




# Q: What is the average daily activity pattern?

Sum of steps by the interval
```{r}
stepsByInterval <- aggregate(steps ~ interval, updatedData, mean)
```

Graphing time series of average steps taken (averaged across all days) by 5-minute intervals
```{r, echo=TRUE}
with(stepsByInterval, 
     plot(interval, 
          steps, 
          type = "l", 
          main = "Avg number of steps over 5-min intervals",
          ylab = "Avg steps taken (averaged across all days)",
          xlab = "5-minute Intervals"))
```

Identifying 5-minute interval where the most steps (on average) occur
```{r}
stepsByInterval[which(stepsByInterval$steps == max(stepsByInterval$steps)),]
```




# Imputing missing values
Calculating the number of rows with NA values in the 'steps' column
```{r}
sum(is.na(rawData$steps))
```

Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
completeData <- rawData
```

Creating loop to fill in NA values with the average value of its specific interval value
```{r}
for (i in 1:nrow(completeData)) {
    if (is.na(completeData[i, 1]) == TRUE) {
        rowInterval <- completeData[i, 3]
        dataSubset <- subset(completeData, interval == rowInterval)
        completeData[i, 1] <- round(mean(dataSubset$steps, na.rm = TRUE))
    }
}
```

Summing steps by the dates
```{r}
stepsByDateComplete <- aggregate(steps ~ date, completeData, sum)
```

Create histogram of steps taken per day
```{r, echo=TRUE}
hist(stepsByDateComplete$steps, 
     main = "Histogram of total number of steps per day",
     xlab = "Steps per day")
```




# Are there differences in activity patterns between weekdays and weekends?
Changing date column from 'factor' to 'date' (POSIXt)
```{r}
completeData$date <- strptime(completeData$date, format = "%Y-%m-%d")
```

Creating list of possible weekdays
```{r}
weekdaysList <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
```

Adding column to dataset that will list if date is a weekday/weekend (initially setting value to NA)
```{r}
completeData$day <- NA
```

Creating loop to evaluate row date and add corresponding weekday/weekend value
```{r}
for (i in 1:nrow(completeData)) {
    rowDate <- completeData[i, 2]
    if (weekdays(rowDate) %in% weekdaysList) {
        completeData[i, 4] <- "weekday"
    } 
    else {
        completeData[i, 4] <- "weekend"
    }
}
```

Loading lattice package
```{r}
library(lattice)
```

Creating subsets of the complete data, weekday vs weekend
```{r}
weekdayData <- subset(completeData, day == "weekday")
weekendData <- subset(completeData, day == "weekend")
stepsByIntervalWeekday <- aggregate(steps ~ interval , weekdayData, mean)
stepsByIntervalWeekday$day <- "weekday"
stepsByIntervalWeekend <- aggregate(steps ~ interval , weekendData, mean)
stepsByIntervalWeekend$day <- "weekend"
```

Combining weekend/weekday data back together
```{r}
stepsByIntervalDays <- rbind(stepsByIntervalWeekday, stepsByIntervalWeekend)
```

Plotting weekend and weekday data
```{r, echo=TRUE}
xyplot(steps ~ interval | day, 
       data = stepsByIntervalDays, 
       type = "l", 
       ylab = "Number of Steps", 
       xlab = "Interval", 
       ylim = c(0, 250), 
       layout = c(1, 2))
```




